Generalizations of Quotient Inductive-Inductive Types\thanks{This work was supported by EFOP-3.6.3-VEKOP-16-2017-00002 grant and COST Action EUTypes CA15123.}

### 2019-10-30 draft/paper.tex
stuff

\todo[inline]{Citations}

@@

The main motivation of the current paper is to extend QIITs so that it
accommodates all algebraic notions of type theories which have been used in
previous works. As a beneficial side effect of fulfilling this goal, infinitary
QIITs such as Cauchy real numbers will be covered as well.

We generalize previous notions of QIITs in the following ways:
\begin{enumerate}
  \item
  \textbf{Large constructors, large elimination} and models at different
  universe levels. This feature is routinely used in the metatheory of type
  theory, but it has not been presented explicitly in previous works about
  QIITs. In order to handle universe levels gracefully, we work in an
  extensional type theory extended with cumulative universes.
  \item
  \textbf{Infinitary constructors}. This enables defining Cauchy real numbers
  and surreal numbers as QIITs. Also, the theory of QIIT signatures is itself
  large and infinitary, thus it can ``eat itself'', i.e.\ include its own
  signature and provide its own metatheory.  We use this self-representation to
  bootstrap the model theory of signatures, without having to assume any
  pre-existing internal syntax.
  \item
  \textbf{Recursive equations}, i.e. equations appearing as assumptions
  of constructors. These have occurred previously in syntaxes of cubical type
  theories, as boundary conditions.
  \item
  \textbf{Sort equations}, or equations between type constructors. They have
  been used recently to conveniently represent type-theoretic universes, for
  example Russell-style universes or cumulative hierarchies.
\end{enumerate}

@@

In the following sections, we will consider QIIT algebras at arbitrary finite
levels, along with large eliminations, where the initial algebra can be at a
different level than the target algebra. For a simple example, consider natural
number algebras at a given level $i$, given as the $\Sigma$-type $NatAlg_i
:\equiv(Nat : \Set_i)\times Nat\times(Nat \ra Nat)$. The initial such algebra is
the set of natural numbers, which is at level $0$, but in type theory we often
want to eliminate into larger $\Set_i$, for example when computing a
$Nat$-indexed family of types. To support convenient reasoning about levels in
this paper, we need two features: cumulativity and a way to quantify over finite
levels.

@@

We need cumulativity to reduce administrative overhead. In the most basic form,
cumulativity requires that whenever $A : \Set_i$, and $i < j$ then also $A :
\Set_j$. We also need cumulativity for $\Sigma$ and function types, in order to
have cumulativity for algebras in general. For example, we want to have $\gamma
: NatAlg_j$ whenever $\gamma : NatAlg_i$ and $i < j$.

@@

From now on, we refer to this theory as cETT (cumulative extensional type
theory). When working in cETT, we use the following notation. We
have Russell-style universes $\Set_i$ indexed by natural numbers, dependent
functions as $(x : A)\ra B$, and dependent pairs as $(x : A)\times B$ with
projections $\proj1$ and $\proj2$. We also use a field projection notation
for iterated pairs. For example, for $t : (A : \Set_i) \times (B : \Set_i)
\times (f : A \ra B)$, we use $t.B$ to denote the projection of the second
component. Sometimes we omit the $t.$ prefix if it is clear from context.

We write propositional equality as $t = u$, with $\refl_t$ for reflexivity. We
have equality reflection for $\blank=\blank$. The unit type is $\top$, with
inhabitant $\tt$.

We also have a lifting operation on types, which introduces cumulativity.
\vspace{-1.5em}
\begin{figure}
\begin{mathpar}
  \inferrule*[lab=universe formation]
             {\\}
             {\Gamma \vdash \Set_i : \Set_{i+j+1}}

@@

In Figure \ref{lifting}, we include an excerpt of cETT's rules. We change
presentation slightly from Sterling. We lift from $i$ to $i+j+1$ levels instead
of assuming $i < j$ and lifting to $j$, and we use a $\U_{\max(i,j)}$ return
type in functions and pairs, instead of having both $A$, $B$ types in the same
$\Set_i$. These changed rules are all derivable from the original ones. In
general, we can expect that lifting never impedes constructions, because it
appropriately computes out of the way.

@@

\begin{nidefinition}{Cumulative subtyping.}
  For $A : \Set_i$ and $B : \Set_{i+j+1}$, we define $A \prec B : \Set_{i+j+1}$
  as $B =\,\,\Lift_j A$.  It follows from term lifting and equality reflection
  that whenever $A \prec B$ and $t : A$, then also $t : B$.
\end{nidefinition}

@@

We also need to quantify over universe levels. cETT does not support this, and
we leave it like that. Instead, for the sake of simplicity, we quantify over
levels in an unspecified metatheory \emph{outside} cETT. Hence, a universe
polymorphic cETT term is understood as a $\mathbb{N}$-indexed family of
terms. We use the $\forall$ symbol for this, e.g.\ as $t : \forall\,i.\,\Set_i \ra
\Set_i$. This is sufficient in this paper, because we do not need to internalize
level-polymorphic constructions in cETT.

@@

\begin{nidefinition}{Notion of model for ToS.}\label{def:tos}
For levels $i$ and $j$, there is a cETT type $\ToS_{i,j} : \Set_{\max(i,\,j)+1}$, whose elements are
ToS models (or ToS-algebras). $\ToS_{i,j}$ is an iterated $\Sigma$-type, containing
all of the following components.
\begin{enumerate}
\item
  A category with families (CwF), where all four underlying sets (of
  objects, morphisms, types and terms) are in $\Set_j$. Following notation in
  \cite{TODO}, we denote these respectively as $\Con : \Set_j$, $\Sub : \Con \ra
  \Con \ra \Set_j$, $\Ty : \Con \ra \Set_j$ and $\Tm : (\Gamma : \Con) \ra
  \Ty\,\Gamma \ra \Set_j$. We denote the empty context as $\emptycon : \Con$,
  context extension as $\blank\ext\blank : (\Gamma : \Con)\ra \Ty\,\Gamma \ra \Con$.
  Substitution on types and terms is written as $\blank[\blank]$.
\item
  A universe $\U : \Ty\,\Gamma$ with decoding $\El : (a : \Tm\,\Gamma\,\U) \ra
  \Ty\,\Gamma$.
\item
  Inductive function space, with $\Pii : (a : \Tm\,\Gamma\,\U) \ra
  \Ty\,(\Gamma\,\ext\,\El\,a) \ra \Ty\,\Gamma$ and application as $\appi :
  \Tm\,\Gamma\,(\Pii\,a\,B)\ra \Tm\,(\Gamma\,\ext\,\El\,a)\,B$. We do not
  support $\lami$ here and in other function types, as they are not essential
  for QIIT signatures.
\item
  External function space with $\Set_i$ domain: $\Pie : (A : \Set_i)\ra(A \ra
  \Ty\,\Gamma)\ra \Ty\,\Gamma$, with $\appe : \Tm\,\Gamma\,(\Pie\,A\,B)\ra(x :
  A)\ra \Tm\,\Gamma\,(B\,x)$.
\item
  Infinitary function space, with $\Piinf : (A : \Set_i)\ra(A \ra
  \Tm\,\Gamma\,\U)\ra \Tm\,\Gamma\,\U$, $\appinf :
  \Tm\,\Gamma\,(\El\,(\Pie\,A\,b))\ra(x : A)\ra \Tm\,\Gamma\,(\El\,(b\,x))$.
\item
  An identity type $\Id : (a : \Tm\,\Gamma\,\U)\ra
  \Tm\,\Gamma\,(\El\,a)\ra\Tm\,\Gamma\,(\El\,a)\ra \Tm\,\Gamma\,\U$, with
  $\Refl : (T : \Tm\,\Gamma\,(\El\,a))\ra \Id\,a\,t\,t$, equality reflection
  and uniqueness of identity proofs.
\item
  An identity type $\IdU : \Tm\,\Gamma\,\U \ra \Tm\,\Gamma\,\U \ra \Ty\,\Gamma$, with
  evident $\ReflU$, equality reflection and uniqueness of identity proofs.
\end{enumerate}
\end{nidefinition}
In the above listing, we omit most CwF components and equations for substitution and
$\beta\eta$-conversion, but these should be understood to be also part of $\ToS_{i,j}$.

\emph{Notational conventions.} CwF components by default support de Bruijn
indices, which are not easily readable. We use instead a nameful notation for
binders in context extension, $\Pii$ and $\lami$, e.g.\ as $(\emptycon \ext a :
U \ext t : \El\,a)$. We also define a type-theoretic version of $\appi$ for
convenience:
\begin{alignat*}{3}
  & \blank\appitt\blank && :
      \Tm\,\Gamma\,(\Pii\,a\,B)\ra
      (u : \Tm\,\Gamma\,(\El\,a)) \ra \Tm\,\Gamma\,(B[id, u])\\
  & t\appitt u && :\equiv (\appi\,t)[id, u]
\end{alignat*}

\begin{nidefinition}{Notion of signature.}
A QIIT signature at level $i$ is a context in an arbitrary $M : \ToS_{i,j}$
model. We define the type of such signatures as follows:
\[
  \Sig_i :\equiv \forall j.\,(M : \ToS_{i,j})\ra M.Con
\]

Note that $\Sig_i$ is defined as a universe-polymorphic type, so it does not
have a type internally to cETT.

\begin{example}
The signature for natural
numbers is the following, where $\emptycon$, $\blank\ext\blank$, $\U$, $\El$ and
$\Pii$ all refer to components of the $\lambda$-bound $M$:
\begin{alignat*}{3}
  & NatSig : \Sig_0 \\
  & NatSig :\equiv \lambda (M : \ToS_{0,j}).\,(\emptycon\,\ext\, N : \U \,\ext\, zero : \El\,N \ext suc : \Pii\,(n : N)\,(\El\,N))
\end{alignat*}
\end{example}

The $i$ level in $\Sig_i$ specifies the level of the external types appearing in
a signatures, since the $\Pie$ and $\Piinf$ function spaces have domains in $\Set_i$.
In the case of naturals numbers, no external types appear, so the level of $NatSig$
can be chosen as $0$.

@@

\begin{example}{Infinitary constructors}. The universe $\U$ is closed under
the $\Piinf$ function type, which makes it possible to write such functions in
the domains of $\Pii$ types. This allows a signature for trees branching with
arbitrary small sets. This is a signature at level 1, since we have $\Set_0$ as
domain type.
\[
TreeSig :\equiv
\emptycon \ext Tree : \U \ext node : \Pie\,\Set_0\,\lambda A.\,\Pii\,(\Piinf\,A\,\lambda \_.\,Tree)\,(\El\,Tree)
\]
\end{example}

@@

\begin{example}{Recursive equations}. Our definition of $\Id$ returns in $\U$,
which, similarly to the case of infinitary functions, allows us to write equations
in $\Pii$ domains. A minimal example:
\[
RecEqSig :\equiv \emptycon \ext A : \U \ext a : \El\,A \ext f : \Pii\,(x : A)\,(\Pii\,(\Id\,A\,x\,a)\,(\El\,A))
\]
There are more complicated and interesting examples, e.g.\ boundary conditions in
various cubical type theories\cite{TODO}. Note that our $\Id$ allows iterated
equations as well, but these are all made trivial in the semantics, where we
assume uniqueness of identity proofs.
\end{example}

@@

\begin{nidefinition}{Finite limit CwFs.}
For a level $i$, there is a cETT type $\flCwF_i : \Set_{i+1}$, which is an
iterated $\Sigma$-type with the following components:
\begin{enumerate}
  \item A CwF with underlying sets all in $\Set_i$.
  \item $\Sigma$-type $\Sg : (A : \Ty\,\Gamma)\ra \Ty\,(\Gamma \ext A)
    \ra \Ty\,\Gamma$, with usual projections and pairing.
  \item Identity type $\Id : (A : \Ty\,\Gamma)\ra \Tm\,\Gamma\,A\ra
    \Tm\,\Gamma\,A\ra \Ty\,\Gamma$, with equality reflection and uniqueness
    of identity proofs.
  \item Constant families. This includes a type former $\Kfam : \Con \ra
    \Ty\,\Gamma$, where $\Gamma$ is implicitly quantified, and an isomorphism
    between $\Sub\,\Gamma\,\Delta$ and $\Tm\,\Gamma\,(\Kfam\,\Delta)$ which is
    natural in $\Gamma$. The idea is that $\Kfam\,\Delta$ is a representation
    of $\Delta$ as a type in any context. Dybjer and Clairambault called constant
    families ``democracy'' in \cite{TODO}.
\end{enumerate}
\end{nidefinition}

@@

\begin{nidefinition} Notion of induction internally to $C : \flCwF_i$:
\begin{alignat*}{3}
  & Inductive : C.\Con \ra \Set_i \\
  & Inductive\,\Gamma :\equiv (A : C.\Ty\,\Gamma)\ra C.\Tm\,\Gamma\,A
\end{alignat*}
\end{nidefinition}

@@

\begin{theorem}{Equivalence of initiality and induction.}
An object $\Gamma : C.\Con$ supports induction if and only if it is
initial. Moreover, induction and initiality are both proof irrelevant
predicates on objects.
\end{theorem}
\begin{proof} Shown by brief internal flCwF reasoning in \cite{TODO}.\qed
\end{proof}

@@

So, why do we use flCwF instead of lex categories? After all, the former
structure has more components and has more immediate complexity than the
latter. The reason is that flCwFs allow us to define induction in strictly the
same way as one would write it down in type theory, since we have $\Ty$ and
$\Tm$ for a primitive notion of dependent objects and morphisms. In contrast,
dependent objects in lex categories is a derived notion, and the induction
principles we get are what we want only up to isomorphisms. This issue is
perhaps not relevant from a purely categorical perspective, but we are concerned
with eventually implementing QIITs in proof assistants, so we prefer if our
semantics computes strictly.

\begin{nidefinition}{flCwF model for ToS.} We have for each $i$ and $j$ levels
an $M : \ToS_{i, 1+\max(i,j+1)}$ such that $M.\Con = \flCwF_{\max(i,j+1)}$. This
model is an extension to the CwF model of signatures in \cite{TODO}, with
$\Sigma$-types in the semantic CwF, the new type formers in signatures, and the
treatment of universe levels. We summarize the model below.
\begin{itemize}
\item As noted, $M.\Con = \flCwF_{\max(i,j+1)}$. The level $i$ marks the level of
      all external sets in function domains, and $j$ marks the level of all inductive
      sets in an algebra. Hence, every algebra, i.e.\ object of the flCwF, has
      level $\max(i,j+1)$. The bump is only needed for $j$, since algebras only contain
      \emph{elements} of $T : \Set_j$ types, while inductive sets are themselves elements
      of $\Set_i$. For example, $NatAlg_i : \Set_{\max(0, i+1)}$.
\item $\sigma : \Sub\,\Gamma\,\Delta$ substitutions are strictly structure-preserving flCwF
      functors.
    \item $A : \Ty\,\Gamma$ types are displayed flCwFs over $\Gamma$. This is an
      flCwF where every underlying set is indexed over underlying sets in
      $\Gamma$, and it is an extension of the displayed categories of Ahrens and
      Lumsdaine\cite{TODO}.  It contains equivalent data to an flCwF slice
      $\delta : \Sub\,\Delta\,\Gamma$. We prefer it to slices because it yields
      expected notions of induction strictly.
\item $t : \Tm\,\Gamma\,A$ terms are sections of the $A$ displayed flCwF. This can be viewed
      as a dependent functor, which maps structure in $\Gamma$ to corresponding displayed
      structure in $A$, e.g.\ it maps objects to displayed objects, morphisms to displayed
      morphisms, and so on.
\item $\emptycon : \Con$ is the terminal flCwF, whose underlying category is the terminal
      category with one object.

\item $\Gamma\,\ext\,A$ is the total flCwF of the $A$ displayed flCwF, where every underlying
      set is given by packing up indexed underlying sets of $A$ in $\Sigma$-s.

\item $\U : \Ty\,\Gamma$ is a displayed flCwF which is constantly (i.e. it does
      not depend on $\Gamma$) the flCwF of sets in $\Set_j$, whose underlying category
      is the usual category of sets.
\item
  $\Pii\,a\,B : \Ty\,\Gamma$ is a displayed flCwF where objects are
  dependent functions from the set of objects of $a$ to the set of objects of
  $B$. It is possible to build such flCwF because the $a : \Tm\,\Gamma\,\U$
  domain is discrete. A general $\Pii$ type is not possible because of variance
  issues; this is an extension of there being no $\Pii$ types in the category
  of categories\cite{elephant}
\item
  $\Pie\,A\,B : \Ty\,\Gamma$ is the $A$-indexed direct product of displayed flCwFs
  given by $B : A \ra \Ty\,\Gamma$.

\end{itemize}

>>>>

ISSUE. We actually need flCwF pseudomorphisms for Sub in infinitary semantics.
$a : \Tm\,\Gamma\,\U$ is \emph{not} a discrete displayed flCwF, only $\El\,a$ is!
$\Piinf\,A\,b : \Tm\,\Gamma,\U$ does not preserve (displayed) object in $\Gamma$
strictly.

\begin{itemize}
\item New semantics with pseudomorphisms for Sub and Tm. Is it enough to
      have weak preservation of comprehension, and we get all other preservation
      for free, see ``gluing for type theory''?
\item Check term model and eliminators to see if change needed.
\item Check if the isomorphism for K in flCwF can be made strict, i.e.
      $\Sub\,\Gamma\,\Delta = \Tm\,\Gamma\,(\Kfam\,\Delta)$.
\end{itemize}

@@

The main motivation of the current paper is to extend the notion of QIIT so that
it accommodates all algebraic notions of type theories which have been used in
previous works. As a side effect of fulfilling this goal, infinitary QIITs such
as Cauchy real numbers are covered as well.

We generalize previous notions of QIITs in the following ways:
\begin{enumerate}
  \item
  \textbf{Large constructors, large elimination} and models at different
  universe levels. This feature is routinely used in the metatheory of type
  theory, but it has not been presented explicitly in previous works about
  QIITs.
  \item
  \textbf{Infinitary constructors}. This allows infinitely branching
  trees. Also, the theory of QIIT signatures is itself large and infinitary,
  thus now it can ``eat itself'', i.e.\ include its own signature and provide
  its own metatheory. This was previously not possible in \cite{TODO}, where
  only finitary QIITs were described. We use this self-representation to
  bootstrap the model theory of signatures, without having to assume any
  pre-existing internal syntax.
  \item
  \textbf{Recursive equations}, i.e. equations appearing as assumptions
  of constructors. These have occurred previously in syntaxes of cubical type
  theories, as boundary conditions\cite{TODO}.
  \item
  \textbf{Sort equations}, or equations between type constructors. They have
  been used recently to conveniently represent type-theoretic universes, for
  example Russell-style universes\cite{TODO} or cumulative hierarchies\cite{TODO}.
\end{enumerate}

@@

In the following sections, we will consider QIIT algebras at arbitrary finite
levels, along with large eliminations, where the initial algebra can be at a
different level than the target algebra. For a simple example, consider natural
number algebras at level $i$, given as the $\Sigma$-type $NatAlg_i :\equiv(Nat :
\Set_i)\times Nat\times(Nat \ra Nat)$. The initial such algebra is the set of
natural numbers, which is at level $0$, but in type theory we often want to
eliminate into larger $\Set_i$, for example when computing a $Nat$-indexed
family of types. To support convenient reasoning about levels in this paper, we
need two features: cumulativity and a way to quantify over finite levels.

@@

We need cumulativity to reduce bureaucratic overhead. In its most basic form,
cumulativity requires that whenever $A : \Set_i$, and $i < j$ then also $A :
\Set_j$. We also need cumulativity for $\Sigma$ and function types, in order to
have cumulativity for algebras in general. For example, we want to have $\gamma
: NatAlg_j$ whenever $\gamma : NatAlg_i$ and $i < j$. We have found that
non-cumulative algebras induce excessive ``lifting'' noise, which we would like
to avoid.

@@

From now on, we refer to this theory as cETT (cumulative extensional type
theory). When working in cETT, we use the following notation. We have
Russell-style universes $\Set_i$ indexed by natural numbers, dependent functions
as $(x : A)\ra B$, and dependent pairs as $(x : A)\times B$ with projections
$\proj1$ and $\proj2$. We also use subscripts as a field projection notation for
iterated pairs. For example, for $t : (A : \Set_i) \times (B : \Set_i) \times (f
: A \ra B)$, we use $B_t$ to denote the projection of the second
component. Sometimes we omit the subscript if it is clear from context.

We write propositional equality as $t = u$, with $\refl_t$ for reflexivity. We
have equality reflection and uniqueness of identity proofs (UIP) for
$\blank=\blank$. The unit type is $\top$, with inhabitant $\tt$.

We also have a lifting operation on types, which introduces cumulativity.
\vspace{-1.5em}
\begin{figure}
\begin{mathpar}
  \inferrule*[lab=universe formation]
             {i < j}
             {\Gamma \vdash \Set_i : \Set_{j}}

@@

In Figure \ref{lifting}, we include an excerpt of cETT's rules. We change
presentation slightly from Sterling: we use a $\U_{\max(i,j)}$ return type in
functions and pairs, instead of having both $A$ and $B$ types in the same
$\Set_i$. These changed rules are all derivable from the original ones. In
general, we can expect that lifting never impedes constructions, because it
appropriately computes out of the way.

@@

\begin{nidefinition}[Cumulative subtyping]
  For $A : \Set_i$ and $B : \Set_{i+j+1}$, we define $A \prec B : \Set_{i+j+1}$
  as $B =\,\,\Lift_j A$.  It follows from term lifting and equality reflection
  that whenever $A \prec B$ and $t : A$, then also $t : B$.
\end{nidefinition}

@@

We also need to quantify over universe levels. cETT does not support this, and
we leave it like that. Instead, for the sake of simplicity, we quantify over
levels in an unspecified metatheory \emph{outside} cETT. Hence, a universe
polymorphic cETT term is understood as a $\mathbb{N}$-indexed family of
terms. We reuse the notation of cETT functions for level-polymorphism, e.g.\
as in the following function:
\[
\lambda i.\,\Set_i\ra\Set_i : (i : \mathbb{N})\ra \Set_{i+1}
\]
In this paper, we do not need to internalize level-polymorphic constructions,
so this setup is sufficient.

@@

\begin{nidefinition}[Notion of model for the theory of signatures]\label{def:tos}
For levels $i$ and $j$, there is a cETT type $\ToS_{i,j} : \Set_{\max(i,\,j)+1}$, whose elements are
ToS models (or ToS-algebras). $\ToS_{i,j}$ is an iterated $\Sigma$-type, containing
all of the following components.
\begin{enumerate}
\item
  A category with families (CwF), where all four underlying sets (of objects,
  morphisms, types and terms) are in $\Set_j$. Following notation in
  \cite{TODO}, we denote these respectively as $\Con : \Set_j$, $\Sub : \Con \ra
  \Con \ra \Set_j$, $\Ty : \Con \ra \Set_j$ and $\Tm : (\Gamma : \Con) \ra
  \Ty\,\Gamma \ra \Set_j$. We denote the empty context as $\emptycon : \Con$,
  context extension as $\blank\ext\blank : (\Gamma : \Con)\ra \Ty\,\Gamma \ra
  \Con$.  Substitution on types and terms is written as $\blank[\blank]$, with
  $\id$ and $\blank\circ\blank$ for identity and composition, and $\blank,\blank
  : (\sigma : \Sub\,\Gamma\,\Delta)\ra \Tm\,\Gamma\,A[\sigma]\ra
  \Sub\,\Gamma\,(\Delta\ext A)$ for substitution extension.
\item
  A universe $\U : \Ty\,\Gamma$ with decoding $\El : (a : \Tm\,\Gamma\,\U) \ra
  \Ty\,\Gamma$.
\item
  Inductive function space, with $\Pii : (a : \Tm\,\Gamma\,\U) \ra
  \Ty\,(\Gamma\,\ext\,\El\,a) \ra \Ty\,\Gamma$ and application as $\appi :
  \Tm\,\Gamma\,(\Pii\,a\,B)\ra \Tm\,(\Gamma\,\ext\,\El\,a)\,B$ and its
  inverse $\lami$.
\item
  External function space with $\Set_i$ domain: $\Pie : (A : \Set_i)\ra(A \ra
  \Ty\,\Gamma)\ra \Ty\,\Gamma$, with $\appe : \Tm\,\Gamma\,(\Pie\,A\,B)\ra((x :
  A)\ra \Tm\,\Gamma\,(B\,x))$ and its inverse $\lame$.
\item
  Infinitary function space, with $\Piinf : (A : \Set_i)\ra(A \ra
  \Tm\,\Gamma\,\U)\ra \Tm\,\Gamma\,\U$, $\appinf :
  \Tm\,\Gamma\,(\El\,(\Pie\,A\,b))\ra((x : A)\ra \Tm\,\Gamma\,(\El\,(b\,x)))$
  and $\laminf$.
\item
  An identity type $\Id : (a : \Tm\,\Gamma\,\U)\ra
  \Tm\,\Gamma\,(\El\,a)\ra\Tm\,\Gamma\,(\El\,a)\ra \Tm\,\Gamma\,\U$, with
  $\Refl : (T : \Tm\,\Gamma\,(\El\,a))\ra \Id\,a\,t\,t$, equality reflection
  and UIP.
\item
  An identity type $\IdU : \Tm\,\Gamma\,\U \ra \Tm\,\Gamma\,\U \ra \Ty\,\Gamma$,
  with reflexivity as $\ReflU$, equality reflection and UIP.
\end{enumerate}
\end{nidefinition}
In the above listing, we omit most CwF components and equations for substitution and
$\beta\eta$-conversion, but these should be understood to be also part of $\ToS_{i,j}$.

\emph{Notational conventions.} CwF components by default support de Bruijn
indices, which are not easily readable. We use instead a nameful notation for
binders in context extension, $\Pii$ and $\lami$, e.g.\ as $(\emptycon \ext a :
U \ext t : \El\,a)$. We also define a type-theoretic version of $\appi$ for
convenience:
\begin{alignat*}{3}
  & \blank\appitt\blank && :
      \Tm\,\Gamma\,(\Pii\,a\,B)\ra
      (u : \Tm\,\Gamma\,(\El\,a)) \ra \Tm\,\Gamma\,(B[\id, u])\\
  & t\appitt u && :\equiv (\appi\,t)[\id, u]
\end{alignat*}

\begin{nidefinition}[Notion of signature]
A QIIT signature at level $i$ is a context in an arbitrary $M : \ToS_{i,j}$
model. We define the type of such signatures as follows:
\[
  \Sig_i :\equiv (j : \mathbb{N})(M : \ToS_{i,j})\ra M.Con
\]

Here, $i$ refers to the level of external types appearing in the signature, in
the domains of $\Pie$ and $\Piinf$ functions, while the quantified $j$ level is
required to allow interpreting a signature in arbitrary-sized ToS models. Note
that $\Sig_i$ is universe-polymorphic, so it does not have a type internally to
cETT.

\begin{example}{
    Signature for natural numbers. Here, no external types appear, so the level
    of $NatSig$ can be chosen as $0$.}
\begin{alignat*}{3}
  & NatSig : \Sig_0 \\
  & NatSig :\equiv \lambda (j :\mathbb{N})(M : \ToS_{0,j}).\\
  & \hspace{2em}(\emptycon_M\,\ext_M\, (N : \U_M) \,\ext_M\, (zero : \El_M\,N) \ext_M (suc : \Pii_M\,N\,(\El_M\,N)))
\end{alignat*}
\end{example}

@@

\begin{example}{Infinitary constructors}. The universe $\U$ is closed under
the $\Piinf$ function type, which makes it possible to write such functions in
the domains of $\Pii$ types. This allows a signature for trees branching with
arbitrary small sets. This is a signature at level 1, since we have $\Set_0$ as
a $\Pie$ domain type.
\[
TreeSig :\equiv
\emptycon \ext (Tree : \U) \ext (node : \Pie\,\Set_0\,\lambda A.\,\Pii\,(\Piinf\,A\,\lambda \_.\,Tree)\,(\El\,Tree))
\]
\end{example}

@@

\begin{example}{Recursive equations}. Our definition of $\Id$ returns in $\U$,
which, similarly to the case of infinitary functions, allows us to write equations
in $\Pii$ domains. A minimal example:
\[
RecEqSig :\equiv \emptycon \ext (A : \U) \ext (a : \El\,A) \ext (f : \Pii\,(x : A)\,(\Pii\,(\Id\,A\,x\,a)\,(\El\,A)))
\]
More interesting (and complicated) examples are boundary conditions in various
cubical type theories\cite{TODO}. Note that our $\Id$ allows iterated equations
as well, but these are all made trivial in the semantics, where we assume UIP.
\end{example}

@@

\begin{nidefinition}[Finite limit CwFs]\label{def:flCwF}
For a level $i$, there is a cETT type $\flCwF_i : \Set_{i+1}$, which is an
iterated $\Sigma$-type with the following components:
\begin{enumerate}
  \item A CwF with underlying sets all in $\Set_i$.
  \item $\Sigma$-type $\Sg : (A : \Ty\,\Gamma)\ra \Ty\,(\Gamma \ext A)
    \ra \Ty\,\Gamma$, with usual projections and pairing.
  \item Identity type $\Id : (A : \Ty\,\Gamma)\ra \Tm\,\Gamma\,A\ra
    \Tm\,\Gamma\,A\ra \Ty\,\Gamma$, with equality reflection and UIP.
  \item Constant families. This includes a type former $\Kfam : \Con \ra
    \Ty\,\Gamma$, where $\Gamma$ is implicitly quantified, and an isomorphism
    between $\Sub\,\Gamma\,\Delta$ and $\Tm\,\Gamma\,(\Kfam\,\Delta)$ which is
    natural in $\Gamma$. The idea is that $\Kfam\,\Delta$ is a representation
    of $\Delta$ as a type in any context. Dybjer and Clairambault called constant
    families ``democracy'' in \cite{TODO}.
\end{enumerate}
\end{nidefinition}

@@

\begin{nidefinition}[Notion of induction in an flCwF]
Given $C : \flCwF_i$, we have the following predicate on objects:
\begin{alignat*}{3}
  & Inductive : \Con_C \ra \Set_i \\
  & Inductive\,\Gamma :\equiv (A : \Ty_C\,\Gamma)\ra \Tm_C\,\Gamma\,A
\end{alignat*}
\end{nidefinition}

@@

\begin{theorem}[Equivalence of initiality and induction]
An object $\Gamma : \Con_C$ supports induction if and only if it is
initial. Moreover, induction and initiality are both proof irrelevant
predicates.
\end{theorem}
\begin{proof} Shown by brief internal flCwF reasoning in \cite{TODO}.\qed
\end{proof}

@@

So, why do we use flCwFs instead of lex categories? After all, the former has
more components and has more immediate complexity than the latter. The reason is
that flCwFs allow us to define induction in strictly the same way as one would
write it down in type theory, since we have $\Ty$ and $\Tm$ for a primitive
notion of dependent objects and morphisms. In contrast, dependent objects in lex
categories is a derived notion, and the induction principles we get are what we
want only up to isomorphisms. This issue is perhaps not relevant from a purely
categorical perspective, but we are concerned with eventually implementing QIITs
in proof assistants, so we prefer if our semantics computes strictly.

\begin{nidefinition}[flCwF model for ToS] We have for each $i$ and $j$ levels
an $M : \ToS_{i, 1+\max(i,j+1)}$ such that $\Con_M = \flCwF_{\max(i,j+1)}$. This
model is an extension of the CwF model of signatures in \cite{TODO}. The new
additions are: $\Sigma$-types in the semantic CwF, the new type formers in
signatures, and the treatment of universe levels. We summarize the model below.

As noted, $\Con$ is given as $\flCwF_{\max(i,j+1)}$. The level $i$ marks the
level of all external sets in function domains, and $j$ marks the level of all
internal sorts in an algebra. Hence, every algebra, i.e.\ object of the flCwF,
has level $\max(i,j+1)$. The bump is only needed for $j$, since algebras only
contain elements of $T : \Set_j$ types, while inductive sets are themselves
elements of $\Set_i$. For example, $NatAlg_i : \Set_{\max(0, i+1)}$.

$\sigma : \Sub\,\Gamma\,\Delta$ substitutions are functors preserving all flCwF
structure strictly.

$A : \Ty\,\Gamma$ types are displayed flCwFs over $\Gamma$. This is an
flCwF where every underlying set is indexed over underlying sets in
$\Gamma$, and it is an extension of the displayed categories of Ahrens and
Lumsdaine\cite{TODO}.  It contains equivalent data to an flCwF slice
$\delta : \Sub\,\Delta\,\Gamma$. We prefer it to slices because it yields
expected notions of induction strictly.

$t : \Tm\,\Gamma\,A$ terms are sections of the displayed flCwF $A$. This can
be viewed as a dependent functor, which maps structure in $\Gamma$ to
corresponding displayed structure in $A$, e.g.\ it maps objects to displayed
objects, morphisms to displayed morphisms, and so on.

$\emptycon : \Con$ is the terminal flCwF, whose underlying category is the
terminal category with one object.

$\Gamma\,\ext\,A$ is the total flCwF of the $A$ displayed flCwF, where every
underlying set is given by packing up indexed underlying sets of $A$ in
$\Sigma$-s.

$\U : \Ty\,\Gamma$ is a displayed flCwF which is constantly (i.e. it does not
depend on $\Gamma$) the flCwF of sets in $\Set_j$, whose underlying category
is the usual category of sets, and which has evident finite limit structure
($\Sg$, $\Id$, $\K$ from Definition \ref{def:flCwF}). Hence, elements of
$\Tm\,\Gamma\,\U$ are functors from $\Gamma$ to \textbf{Set}. $\El :
\Tm\,\Gamma\,\U\ra\Ty\,\Gamma$ interprets such functors as discrete displayed
flCwFs, where all morphisms are identities.

$\Pii\,a\,B : \Ty\,\Gamma$ is a displayed flCwF where objects are dependent
functions from the set of objects of $a$ to the set of objects of $B$, and
morphisms are pointwise morphisms. This is possible to build because the $a :
\Tm\,\Gamma\,\U$ domain is discrete. A general $\Pii$ type is not possible
because of variance issues; this follows from the fact that the category of
categories does not support $\Pii$ types\cite{elephant}. Finite limit structure
is given pointwise, and is inherited from the domain $B$.

$\Pie\,A\,B : \Ty\,\Gamma$ is the $A$-indexed direct product of the family of
displayed flCwFs given by $B : A \ra \Ty\,\Gamma$.

$\Piinf\,A\,b : \Tm\,\Gamma\,\U$ is likewise the $A$-indexed direct product of
the family of discrete displayed flCwFs given by $b : A \ra \Tm\,\Gamma\,\U$.

$\Id\,a\,t\,u : \Tm\,\Gamma\,\U$ maps every object $\gamma$ of $\Gamma$ to the
set $t\,\gamma = u\,\gamma$, and every morphism $\sigma :
\Sub_\Gamma\,\gamma\,\gamma'$ to a function with type $t\,\gamma = u\,\gamma
\ra t\,\gamma' = u\,\gamma'$, which can be given by the functoriality of $t$
and $u$. In short, $\Id\,a\,t\,u$ expresses pointwise equality of $t$ and $u$
dependent functors.

$\IdU\,a\,b : \Ty\,\Gamma$ is a displayed flCwF where the set of objects over
$\gamma : \Con_\Gamma$ is $a\,\gamma = b\,\gamma$, that is, a strict equation
of types in $\Set_j$. Displayed morphisms are trivial (i.e.\ constantly
$\top$) because of UIP in cETT.
